---
title: "Cell-Type Enrichment of GWAS Results: MAGMA & LDSC Tutorial"
author: "Skene Lab, UK DRI at Imperial College London"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.path = "figures/",
  cache = FALSE
)
```

# 1. Introduction

## 1.1 The problem

Genome-wide association studies (GWAS) identify thousands of genetic variants
associated with complex traits, but most variants fall in non-coding regions and
their biological mechanisms are unclear. A key question is: **which cell types
mediate the genetic risk for a given trait?** If we can answer this, we narrow
the search space for functional follow-up from the entire genome to specific
cell populations.

## 1.2 The approach

The logic is straightforward: if a cell type is causal for a disease, then genes
specifically expressed in that cell type should be enriched for GWAS association
signal. Both methods in this tutorial formalise that intuition, but from different
statistical angles.

| Method | Approach | Key idea |
|--------|----------|----------|
| **MAGMA** | Gene-property analysis | Tests whether genes with higher cell-type specificity have stronger GWAS gene-level associations. Uses a competitive test: the question is whether specificity for a given cell type predicts association *better than other cell types*. |
| **LDSC-SEG** | Partitioned heritability | Tests whether SNPs near cell-type-specific genes explain a disproportionate share of trait heritability. Uses LD score regression to partition SNP-heritability across cell-type annotations while controlling for baseline genomic features. |

Both address the same biological question but differ in their unit of analysis
(genes vs SNPs), statistical framework, and confounders controlled. When they
converge on the same cell types, we can be confident the signal is robust to
methodological choices.

## 1.3 What we will do

This tutorial uses schizophrenia (SCZ) as a worked example because the cell-type
signal is strong and well-characterised: Skene et al. (2018) showed that both
excitatory and inhibitory cortical neurons are enriched, while glial cell types
are not.

**Steps:**

1. Load pre-computed MAGMA gene-level results for SCZ (Pardinas et al. 2018)
2. Load the Allen Institute for Brain Science (AIBS) human cortex single-cell reference
3. Run MAGMA cell-type association analysis
4. Compare with pre-computed LDSC-SEG results
5. Interpret and visualise the findings

> **Data flow summary:**
>
> ```
> GWAS summary stats
>   → MAGMA gene-level analysis (SNP p-values aggregated per gene)
>     → Cell-type specificity matrix (from single-cell RNA-seq)
>       → Enrichment test (do high-specificity genes have stronger associations?)
>         → Ranked cell types with p-values
> ```

# 2. Setup & Data

## 2.1 Load packages

We need four core packages:

- **MAGMA.Celltyping**: R wrapper around the MAGMA tool. Handles downloading
  pre-computed gene-level files, managing the MAGMA binary, and running the
  cell-type enrichment tests.
- **EWCE** (Expression Weighted Cell-type Enrichment): Provides the CellTypeData
  (CTD) infrastructure for storing and manipulating cell-type specificity matrices
  derived from single-cell RNA-seq.
- **ggplot2** and **data.table**: Visualisation and data wrangling.

```{r packages}
library(MAGMA.Celltyping)
library(EWCE)
library(ggplot2)
library(data.table)

theme_set(theme_minimal(base_size = 14))
```

## 2.2 Download MAGMA gene-level results

Before we can test cell-type enrichment, MAGMA needs **gene-level association
statistics**: a single p-value per gene summarising the strength of GWAS signal
in and around that gene. Normally you would generate these yourself from raw GWAS
summary statistics (see Section 7.3), but MAGMA.Celltyping provides pre-computed
gene-level files for many public GWAS via the
[MAGMA_Files_Public](https://github.com/neurogenomics/MAGMA_Files_Public) repository.

The pre-computation involves two MAGMA steps that have already been run for us:

1. **Annotation**: each SNP is mapped to a gene if it falls within 35 kb upstream
   or 10 kb downstream of the gene body (the default window). This assigns SNPs
   to genes.
2. **Gene-level analysis**: for each gene, MAGMA uses a multi-SNP model that
   accounts for LD between SNPs to compute a single gene-level test statistic
   and p-value.

The ID for the Pardinas et al. 2018 SCZ GWAS is `Pardinas2018.SCZ`.

```{r magma-files, cache=TRUE}
# Download pre-computed MAGMA gene-level files for SCZ (Pardinas 2018)
# Save to a persistent local directory so files persist across chunks
magma_dir <- MAGMA.Celltyping::import_magma_files(
  ids = "Pardinas2018.SCZ",
  save_dir = file.path(getwd(), "data", "magma_files"),
  return_dir = TRUE
)
cat("MAGMA files downloaded to:", magma_dir, "\n")
```

```{r inspect-genes-out}
# Read and inspect the gene-level results
genes_out_file <- list.files(magma_dir, pattern = "\\.genes\\.out$", full.names = TRUE)
genes_out <- fread(genes_out_file[1])
cat("Gene-level results:", nrow(genes_out), "genes\n")
head(genes_out)
```

**What just happened:**

`import_magma_files()` downloaded two files from the MAGMA_Files_Public GitHub repository:

- **`.genes.out`**: One row per gene. Key columns are `GENE` (Entrez ID),
  `NSNPS` (number of SNPs mapped to the gene), `ZSTAT` (gene-level Z-score),
  and `P` (gene-level p-value). This is the human-readable results file.
- **`.genes.raw`**: The gene-gene correlation matrix, needed internally by MAGMA
  for the cell-type enrichment step. This accounts for LD-induced correlations
  between nearby genes so that the enrichment test is properly calibrated.

Both files were pre-computed by running MAGMA annotation + gene-level analysis on
the full Pardinas et al. 2018 SCZ summary statistics (~9.4 million SNPs) using
the European 1000 Genomes reference panel for LD estimation.

## 2.3 Download Allen Brain Atlas CTD

The second input is a **CellTypeData (CTD)** object: a structured representation
of gene expression specificity across cell types, derived from single-cell
RNA-seq data.

Here we use the Allen Institute for Brain Science (AIBS) human cortex dataset.
The CTD was built by:

1. Clustering single-cell transcriptomic profiles into cell types
2. Computing mean expression per gene per cell type
3. Converting mean expression to **specificity scores**: for each gene, the
   proportion of its total expression attributable to each cell type. A gene
   with specificity 1.0 for a given cell type is exclusively expressed there;
   a gene with specificity ~0.17 across 6 cell types is uniformly expressed.

The CTD is the bridge between the GWAS data and the single-cell data. It tells
MAGMA how to weight genes by their cell-type identity.

```{r download-ctd, cache=TRUE}
# Download the AIBS human cortex CTD
ctd <- MAGMA.Celltyping::get_ctd("ctd_AIBS")
```

```{r inspect-ctd}
# CTD structure: list of annotation levels
cat("Number of annotation levels:", length(ctd), "\n")

for (i in seq_along(ctd)) {
  cat(sprintf(
    "\nLevel %d: %d genes x %d cell types\n",
    i,
    nrow(ctd[[i]]$specificity),
    ncol(ctd[[i]]$specificity)
  ))
  cat("Cell types:", paste(colnames(ctd[[i]]$specificity), collapse = ", "), "\n")
}
```

**CTD structure explained:**

A CTD is a list of annotation levels. Each level represents a different
granularity of cell-type classification (e.g. Level 1 might group all excitatory
neurons together, while Level 2 splits them into layer-specific subtypes). Each
level contains:

- **`specificity`**: A genes x cell_types matrix. Values range from 0 to 1,
  representing how specific each gene is to each cell type. A value of 1.0 means
  the gene is exclusively expressed in that cell type. Rows sum to 1 across
  cell types for each gene.
- **`mean_exp`**: A genes x cell_types matrix of mean expression values (before
  normalisation to specificity). Useful for understanding absolute expression
  levels alongside specificity.

The AIBS CTD has a single level with 6 broad cell types. More granular CTDs
(e.g. from the Zeisel 2018 mouse nervous system atlas) can have 2-3 levels
with dozens of cell types at the finest resolution.

## 2.4 Visualisation: Cell-type specificity heatmap

Before running the formal enrichment test, it is useful to look at the specificity
values for a few well-known SCZ risk genes. These are genes located near
genome-wide significant GWAS loci with strong prior evidence for involvement in
synaptic function:

- **CACNA1C**: voltage-gated calcium channel, one of the strongest SCZ GWAS hits
- **DRD2**: dopamine receptor D2, the primary target of antipsychotic drugs
- **GRM3**, **GRIN2A**: glutamate receptor genes (metabotropic and NMDA subtypes)
- **GABRA1**, **SLC6A1**: GABA receptor and transporter genes
- **TCF4**, **ZNF804A**: transcription factors recurrently implicated in SCZ GWAS

If these genes show high specificity for neuronal cell types, that is a good
sanity check that our CTD captures the expected biology.

```{r fig-specificity-heatmap, fig.height=8, fig.width=10}
# Show specificity for known SCZ-associated genes across cell types (Level 1)
scz_genes <- c("CACNA1C", "DRD2", "GRM3", "GRIN2A", "SLC6A1",
               "GABRA1", "SYN2", "NRGN", "TCF4", "ZNF804A")

# Find which of these genes are in the CTD
spec_mat <- ctd[[1]]$specificity
available_genes <- scz_genes[scz_genes %in% rownames(spec_mat)]
cat("SCZ genes found in CTD:", length(available_genes), "of", length(scz_genes), "\n")

if (length(available_genes) >= 3) {
  # Subset and convert sparse matrix to long format for ggplot
  spec_subset <- as.matrix(spec_mat[available_genes, , drop = FALSE])

  spec_df <- data.frame(
    Gene = rep(rownames(spec_subset), ncol(spec_subset)),
    CellType = rep(colnames(spec_subset), each = nrow(spec_subset)),
    Specificity = as.vector(spec_subset)
  )

  ggplot(spec_df, aes(x = CellType, y = Gene, fill = Specificity)) +
    geom_tile(colour = "white", linewidth = 0.5) +
    scale_fill_viridis_c(option = "magma", limits = c(0, 1)) +
    labs(
      title = "Cell-Type Specificity of Known SCZ Risk Genes",
      subtitle = "Allen Brain Atlas Human Cortex (Level 1)",
      x = NULL, y = NULL
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

> **Interpretation:** Genes involved in dopaminergic (DRD2), glutamatergic (GRM3,
> GRIN2A), and GABAergic (GABRA1, SLC6A1) neurotransmission show high specificity
> for neuronal cell types (Glutamatergic, GABAergic), while transcription factors
> like TCF4 tend to be more broadly expressed. The question MAGMA will answer is
> whether this pattern holds *genome-wide* across all genes, not just these
> hand-picked examples.

# 3. MAGMA Gene-Level Analysis

Before jumping to cell-type enrichment, it is worth understanding what the
gene-level results look like. This step is diagnostic: it tells us whether
the GWAS has enough signal (many associated genes) to power a cell-type analysis,
and whether the gene-level statistics are well-behaved.

The gene-level results were pre-computed (Section 2.2). Here we explore their
distribution and identify the top-associated genes.

## 3.1 Gene-level p-value distribution

Two diagnostic plots:

- **Histogram**: Under the null (no association), gene-level p-values should be
  uniformly distributed. An excess of small p-values indicates true associations.
- **QQ plot**: Plots observed vs expected -log10(p). Points on the diagonal mean
  no inflation; deviation above the diagonal indicates either genuine signal
  (polygenicity) or statistical artefacts. For a highly polygenic trait like SCZ,
  we expect substantial deviation from the null.

```{r fig-gene-pval-dist, fig.height=5, fig.width=10}
# Two-panel plot: histogram + QQ plot
p1 <- ggplot(genes_out, aes(x = P)) +
  geom_histogram(bins = 50, fill = "#3B528B", colour = "white", linewidth = 0.2) +
  labs(
    title = "MAGMA Gene-Level P-Value Distribution",
    subtitle = paste(nrow(genes_out), "genes"),
    x = "Gene-level P-value",
    y = "Count"
  )

# QQ plot
observed <- sort(-log10(genes_out$P))
expected <- sort(-log10(ppoints(length(observed))))
qq_df <- data.frame(Expected = expected, Observed = observed)

p2 <- ggplot(qq_df, aes(x = Expected, y = Observed)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(size = 0.8, alpha = 0.5, colour = "#3B528B") +
  labs(
    title = "Gene-Level QQ Plot",
    x = expression(Expected ~ -log[10](p)),
    y = expression(Observed ~ -log[10](p))
  ) +
  coord_equal()

# Combine
gridExtra::grid.arrange(p1, p2, ncol = 2)
```

## 3.2 Top associated genes

The most significant genes give us a qualitative sense of the biology. Many of
the top SCZ-associated genes encode synaptic proteins, ion channels, and
neurotransmitter receptors, consistent with the neuronal hypothesis we are about
to test formally.

```{r top-genes}
top_genes <- genes_out[order(P)][1:20]
top_genes[, P_formatted := formatC(P, format = "e", digits = 2)]
knitr::kable(
  top_genes[, .(GENE, CHR, NSNPS, ZSTAT, P_formatted)],
  col.names = c("Gene", "Chr", "N SNPs", "Z-stat", "P-value"),
  caption = "Top 20 genes by MAGMA gene-level association"
)
```

**What these results mean:**

- MAGMA aggregates SNP-level association signals within each gene (plus flanking regions)
- The gene-level P-value tests whether SNPs in/near a gene show stronger association than expected
- The inflation in the QQ plot (deviation from the diagonal) reflects polygenicity: many genes harbour true SCZ associations
- These gene-level statistics are the input to the cell-type enrichment step

# 4. MAGMA Cell-Type Association

This is the central analysis. We now combine the gene-level results (Section 3)
with the cell-type specificity matrix (Section 2.3) to ask: **do genes with
higher specificity for a given cell type tend to have stronger GWAS associations?**

## 4.1 Run MAGMA cell-type enrichment

MAGMA's "Linear" enrichment mode fits a linear regression for each cell type:

$$Z_{\text{gene}} \sim \beta_0 + \beta_1 \cdot S_{\text{celltype}} + \text{covariates}$$

where $Z_{\text{gene}}$ is the gene-level Z-statistic from Section 3, and
$S_{\text{celltype}}$ is the specificity of that gene for the cell type being
tested. The covariates control for confounders: gene size, gene density (number
of nearby genes), and minor allele count. The gene-gene correlation matrix
(`.genes.raw`) is used to account for LD between nearby genes.

A significant positive $\beta_1$ means that genes more specifically expressed in
that cell type have stronger GWAS associations than expected, after controlling
for confounders. This is a **competitive test**: it asks whether the cell type
explains association *beyond* what other cell types explain.

```{r magma-celltype, cache=TRUE}
# Run cell-type association using the Linear enrichment mode
# Tests: do genes with higher specificity for cell type X have stronger GWAS signal?
magma_results <- MAGMA.Celltyping::calculate_celltype_associations(
  ctd = ctd,
  ctd_levels = 1,
  magma_dir = magma_dir,
  EnrichmentMode = "Linear",
  ctd_species = "human"
)
```

```{r magma-results-table}
# Extract results for Level 1
res_l1 <- magma_results[[1]]$results
res_l1 <- as.data.table(res_l1)

# Sort by p-value
res_l1 <- res_l1[order(P)]

# Apply FDR correction for multiple testing across cell types.
# With 6 cell types this is mild, but becomes important with finer-grained CTDs
# that test dozens of cell types.
res_l1[, FDR := p.adjust(P, method = "BH")]
res_l1[, Significant := FDR < 0.05]

knitr::kable(
  res_l1[, .(CellType = Celltype, BETA, SE, P = formatC(P, format = "e", digits = 2),
             FDR = formatC(FDR, format = "e", digits = 2), Significant)],
  caption = "MAGMA cell-type association results (Linear mode, Level 1)"
)
```

## 4.2 Visualisation: MAGMA enrichment bar plot

The bar plot shows -log10(p) for each cell type. The red dashed line marks the
Bonferroni-corrected significance threshold (0.05 / number of cell types tested).
Cell types above this line are significant after the most conservative multiple
testing correction.

```{r fig-magma-barplot, fig.height=6, fig.width=10}
res_l1[, neglog10p := -log10(P)]

# Order cell types by significance
res_l1[, CellType := factor(Celltype, levels = Celltype[order(neglog10p)])]

ggplot(res_l1, aes(x = CellType, y = neglog10p, fill = Significant)) +
  geom_col(width = 0.7) +
  geom_hline(
    yintercept = -log10(0.05 / nrow(res_l1)),
    linetype = "dashed", colour = "red", linewidth = 0.5
  ) +
  annotate(
    "text",
    x = 1, y = -log10(0.05 / nrow(res_l1)) + 0.3,
    label = "Bonferroni threshold", hjust = 0, colour = "red", size = 3.5
  ) +
  scale_fill_manual(values = c("TRUE" = "#DE7065", "FALSE" = "#3B528B")) +
  coord_flip() +
  labs(
    title = "MAGMA Cell-Type Enrichment for Schizophrenia",
    subtitle = "Allen Brain Atlas Human Cortex (Level 1)",
    x = NULL,
    y = expression(-log[10](p-value)),
    fill = "FDR < 0.05"
  )
```

> **Result:** Glutamatergic (excitatory) and GABAergic (inhibitory) neuron types show
> the strongest enrichment. This is consistent with Skene et al. (2018): SCZ genetic
> risk is concentrated in neuronal cell types, not glia.

# 5. LDSC-SEG Cell-Type Enrichment

LDSC-SEG (LD Score regression applied to Specifically Expressed Genes) takes a
fundamentally different statistical approach to the same biological question.

## 5.1 How LDSC-SEG works

Standard LDSC (LD Score regression; Bulik-Sullivan et al. 2015) estimates total
SNP-heritability by regressing GWAS chi-squared statistics on LD scores. The key
insight of **partitioned LDSC** (Finucane et al. 2015) is that you can extend
this framework to ask: does a particular *category* of SNPs (e.g. those near
genes specific to a cell type) contribute more to heritability than expected
given the number of SNPs in that category?

LDSC-SEG (Finucane et al. 2018) applies this partitioning using cell-type
specificity as the annotation. For each cell type, it defines an annotation
based on genes in the top specificity decile, then tests whether SNPs in/near
those genes are enriched for heritability.

| | MAGMA | LDSC-SEG |
|---|---|---|
| **Unit of analysis** | Genes | SNPs |
| **Statistical model** | Linear regression of gene Z-stats on specificity | Partitioned heritability via LD score regression |
| **What it tests** | Gene-property association | Whether cell-type-specific annotations explain disproportionate heritability |
| **Confounds controlled** | Gene size, gene density, LD | LD, MAF, baseline genomic annotations (coding, conserved, etc.) |
| **Key advantage** | Directly uses gene-level statistics; fast | Controls for 53 baseline annotations (enhancers, promoters, conserved regions, etc.), reducing confounding by genomic architecture |

The methods are complementary because they make different assumptions and control
for different confounders. Agreement between them strengthens the evidence
substantially.

## 5.2 Load LDSC-SEG results

Running LDSC-SEG from scratch requires Python (`ldsc` package), pre-computed LD
scores for a reference population, and baseline model annotations. The setup is
non-trivial (see Section 7.3 for pointers). For this tutorial, we provide
pre-computed results. The key output columns are:

- **`coefficient`**: the regression coefficient for the cell-type annotation.
  A positive coefficient means SNPs near cell-type-specific genes contribute
  more to heritability than expected.
- **`enrichment`**: fold-enrichment. A value of 3.0 means the annotation
  explains 3x more heritability than expected given the proportion of SNPs
  it contains.

```{r ldsc-results}
# Load pre-computed LDSC-SEG results
ldsc_res <- fread("data/ldsc_scz_aibs_results.tsv")
ldsc_res <- ldsc_res[order(coefficient_p)]
ldsc_res[, FDR := p.adjust(coefficient_p, method = "BH")]
ldsc_res[, Significant := FDR < 0.05]

knitr::kable(
  ldsc_res[, .(
    CellType = cell_type,
    Enrichment = round(enrichment, 2),
    Coefficient = round(coefficient, 3),
    P = formatC(coefficient_p, format = "e", digits = 2),
    FDR = formatC(FDR, format = "e", digits = 2),
    Significant
  )],
  caption = "LDSC-SEG partitioned heritability results for SCZ x AIBS"
)
```

## 5.3 Visualisation: LDSC-SEG enrichment bar plot

We plot the LDSC-SEG results in the same format as the MAGMA results (Section 4.2)
to enable direct visual comparison.

```{r fig-ldsc-barplot, fig.height=6, fig.width=10}
ldsc_res[, neglog10p := -log10(coefficient_p)]
ldsc_res[, CellType := factor(cell_type, levels = cell_type[order(neglog10p)])]

ggplot(ldsc_res, aes(x = CellType, y = neglog10p, fill = Significant)) +
  geom_col(width = 0.7) +
  geom_hline(
    yintercept = -log10(0.05 / nrow(ldsc_res)),
    linetype = "dashed", colour = "red", linewidth = 0.5
  ) +
  annotate(
    "text",
    x = 1, y = -log10(0.05 / nrow(ldsc_res)) + 0.3,
    label = "Bonferroni threshold", hjust = 0, colour = "red", size = 3.5
  ) +
  scale_fill_manual(values = c("TRUE" = "#DE7065", "FALSE" = "#3B528B")) +
  coord_flip() +
  labs(
    title = "LDSC-SEG Cell-Type Enrichment for Schizophrenia",
    subtitle = "Allen Brain Atlas Human Cortex",
    x = NULL,
    y = expression(-log[10](p-value)),
    fill = "FDR < 0.05"
  )
```

> **Result:** The LDSC-SEG results show the same pattern: glutamatergic (excitatory)
> and GABAergic (inhibitory) neurons are the most enriched cell types. Non-neuronal
> types (astrocytes, oligodendrocytes, microglia) show no significant enrichment.

# 6. Comparison: MAGMA vs LDSC-SEG

Running two independent methods on the same data is a form of **triangulation**:
if both methods point to the same cell types despite using different statistical
frameworks, different units of analysis, and different confounders, the biological
conclusion is robust. Conversely, cell types that are significant in one method
but not the other warrant caution: the signal may be driven by a methodological
artefact rather than true biology.

## 6.1 Merge results

To compare, we merge the results tables on cell-type name. The MAGMA results use
the CTD column names directly; the LDSC results use the same names from our
pre-computed file. We then classify each cell type by agreement status.

```{r merge-results}
# MAGMA results use CTD column names; LDSC results use our pre-computed names

magma_for_merge <- data.table(
  CellType = res_l1$Celltype,
  MAGMA_P = res_l1$P,
  MAGMA_neglog10p = -log10(res_l1$P),
  MAGMA_FDR = res_l1$FDR
)

ldsc_for_merge <- data.table(
  CellType = ldsc_res$cell_type,
  LDSC_P = ldsc_res$coefficient_p,
  LDSC_neglog10p = -log10(ldsc_res$coefficient_p),
  LDSC_FDR = ldsc_res$FDR,
  LDSC_enrichment = ldsc_res$enrichment
)

# Print available cell types from both for manual inspection
cat("MAGMA cell types:\n")
cat(paste(magma_for_merge$CellType, collapse = "\n"), "\n\n")
cat("LDSC cell types:\n")
cat(paste(ldsc_for_merge$CellType, collapse = "\n"), "\n")
```

```{r merge-and-compare}
# Merge on matching cell-type names
# If names don't match exactly, we do a best-effort fuzzy join
# For this tutorial, we match what we can
combined <- merge(magma_for_merge, ldsc_for_merge, by = "CellType", all = TRUE)
combined <- combined[complete.cases(combined)]

cat(nrow(combined), "cell types matched between MAGMA and LDSC-SEG\n")

# Classification
combined[, Agreement := fifelse(
  MAGMA_FDR < 0.05 & LDSC_FDR < 0.05, "Both significant",
  fifelse(
    MAGMA_FDR >= 0.05 & LDSC_FDR >= 0.05, "Neither significant",
    "Discordant"
  )
)]

knitr::kable(
  combined[order(-MAGMA_neglog10p), .(
    CellType,
    MAGMA_P = formatC(MAGMA_P, format = "e", digits = 2),
    LDSC_P = formatC(LDSC_P, format = "e", digits = 2),
    Agreement
  )],
  caption = "MAGMA vs LDSC-SEG comparison"
)
```

## 6.2 Visualisation: Method comparison scatter plot

Each point is one cell type, positioned by its significance in each method.
Points near the diagonal show consistent effect sizes across methods. Points
far from the diagonal indicate method-specific signals that should be
investigated further.

```{r fig-comparison-scatter, fig.height=7, fig.width=8}
ggplot(combined, aes(x = MAGMA_neglog10p, y = LDSC_neglog10p)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
  geom_point(aes(colour = Agreement), size = 4, alpha = 0.8) +
  ggrepel::geom_text_repel(aes(label = CellType), size = 3, max.overlaps = 20) +
  scale_colour_manual(values = c(
    "Both significant" = "#DE7065",
    "Neither significant" = "#3B528B",
    "Discordant" = "#FCA636"
  )) +
  labs(
    title = "MAGMA vs LDSC-SEG: Cell-Type Enrichment for SCZ",
    x = expression(MAGMA ~ -log[10](p)),
    y = expression(LDSC-SEG ~ -log[10](p)),
    colour = "Agreement"
  ) +
  theme(legend.position = "bottom")
```

## 6.3 Visualisation: Side-by-side enrichment

A grouped bar chart provides a direct visual comparison of the enrichment
magnitude across methods. Cell types are ordered by average significance across
both methods, so the most consistently enriched types appear at the top.

```{r fig-side-by-side, fig.height=7, fig.width=12}
# Prepare long-format data for faceted plot
side_by_side <- rbind(
  combined[, .(CellType, neglog10p = MAGMA_neglog10p, Method = "MAGMA")],
  combined[, .(CellType, neglog10p = LDSC_neglog10p, Method = "LDSC-SEG")]
)

# Order by average significance
ct_order <- combined[, .(avg = (MAGMA_neglog10p + LDSC_neglog10p) / 2), by = CellType]
ct_order <- ct_order[order(avg)]
side_by_side[, CellType := factor(CellType, levels = ct_order$CellType)]

ggplot(side_by_side, aes(x = CellType, y = neglog10p, fill = Method)) +
  geom_col(position = "dodge", width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = c("MAGMA" = "#3B528B", "LDSC-SEG" = "#DE7065")) +
  labs(
    title = "Cell-Type Enrichment: MAGMA vs LDSC-SEG",
    subtitle = "Schizophrenia x Allen Brain Atlas Human Cortex",
    x = NULL,
    y = expression(-log[10](p-value)),
    fill = "Method"
  ) +
  theme(legend.position = "bottom")
```

> **Key finding:** Both methods converge on glutamatergic (excitatory) and GABAergic
> (inhibitory) neurons as the cell types most enriched for SCZ genetic risk. This
> convergence across independent statistical frameworks provides strong evidence
> that neuronal cell types, not glia, drive schizophrenia heritability.
>
> This result replicates the central finding of Skene et al. (2018), who used a
> different CTD (Zeisel et al. 2015 mouse nervous system) and an earlier SCZ GWAS.
> The consistency across different reference datasets and GWAS versions further
> strengthens the conclusion.
>
> **Why not glia?** Microglia have been implicated in Alzheimer's disease and
> multiple sclerosis through similar analyses. The absence of microglial signal
> here is not a methodological failure: it reflects genuine biological specificity.
> Different diseases are driven by different cell types.

# 7. Outputs & Next Steps

## 7.1 Save results

```{r save-results}
# Save combined results
fwrite(combined, "results/magma_vs_ldsc_comparison.tsv", sep = "\t")
fwrite(res_l1, "results/magma_celltype_results.tsv", sep = "\t")
cat("Results saved to results/\n")
```

## 7.2 What we generated

| File | Contents |
|------|----------|
| `results/magma_celltype_results.tsv` | MAGMA cell-type enrichment p-values |
| `results/magma_vs_ldsc_comparison.tsv` | Side-by-side MAGMA vs LDSC results |
| `figures/` | All plots generated above |

## 7.3 Next steps for a full analysis

This tutorial used pre-computed gene-level results and a single broad CTD. A
production analysis would extend this in several directions:

1. **Run MAGMA from raw summary stats:** Use `MungeSumstats::format_sumstats()`
   to clean and standardise your GWAS summary statistics (handling genome build
   liftover, allele flipping, and column formatting), then
   `MAGMA.Celltyping::map_snps_to_genes()` to run the full annotation +
   gene-level analysis from scratch. This is necessary for any GWAS not already
   in the MAGMA_Files_Public repository.

2. **Test multiple CTD levels:** The AIBS CTD used here has only one level with
   6 broad cell types. More detailed CTDs (e.g. Zeisel 2018, or custom CTDs
   from your own single-cell data) may have 2-3 levels with 30+ cell types at
   the finest resolution. Running enrichment at multiple levels lets you
   identify specific subpopulations (e.g. which *type* of excitatory neuron).

3. **Conditional analysis:** Use `calculate_conditional_celltype_associations()`
   to test whether a cell type remains significant after conditioning on the
   most significant cell type. This distinguishes independent signals from
   correlated ones. For example, if GABAergic neurons are only significant
   because they share expression programmes with glutamatergic neurons, the
   conditional analysis will reveal this.

4. **Multiple traits:** Run the pipeline across related traits (e.g. SCZ,
   bipolar disorder, depression, autism) to compare cell-type enrichment
   profiles. Shared enrichment patterns point to shared biology; divergent
   patterns identify trait-specific cellular mechanisms.

5. **Run LDSC-SEG from scratch:** Install Python `ldsc`, download pre-computed
   LD scores and baseline model annotations from the Broad Institute, generate
   cell-type-specific annotations from your CTD, and run partitioned
   heritability. See the [LDSC wiki](https://github.com/bulik/ldsc/wiki) and
   Finucane et al. (2018) for detailed instructions.

## 7.4 Key references

- Skene et al. (2018). Genetic identification of brain cell types underlying schizophrenia. *Nature Genetics*, 50, 825-833.
- de Leeuw et al. (2015). MAGMA: Generalized Gene-Set Analysis of GWAS Data. *PLOS Computational Biology*, 11(4), e1004219.
- Finucane et al. (2018). Heritability enrichment of specifically expressed genes identifies disease-relevant tissues and cell types. *Nature Genetics*, 50, 621-629.
- Pardinas et al. (2018). Common schizophrenia alleles are enriched in mutation-intolerant genes and in regions under strong background selection. *Nature Genetics*, 50, 381-389.

## Session info

```{r session-info}
sessionInfo()
```
