---
title: "Cell-Type Enrichment of GWAS Results: MAGMA & LDSC Tutorial"
author: "Skene Lab, UK DRI at Imperial College London"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.path = "figures/",
  cache = FALSE
)
```

# 1. Introduction

**Goal:** Given GWAS summary statistics for a trait, identify which brain cell types
are enriched for genetic association signal. This tells us which cell types are
likely driving genetic risk.

We use two complementary methods:

| Method | Approach | Key idea |
|--------|----------|----------|
| **MAGMA** | Gene-property analysis | Tests whether genes with higher cell-type specificity have stronger GWAS associations |
| **LDSC-SEG** | Partitioned heritability | Tests whether SNPs near cell-type-specific genes explain more heritability than expected |

Both methods address the same question from different angles. When they converge on
the same cell types, confidence is high.

**In this tutorial we will:**

1. Load pre-computed MAGMA gene-level results for schizophrenia (Pardinas et al. 2018)
2. Load the Allen Institute for Brain Science (AIBS) human cortex cell-type reference
3. Run MAGMA cell-type association analysis
4. Compare with LDSC-SEG results
5. Interpret and visualise the findings

> **Data flow summary:** GWAS summary stats → MAGMA gene-level → cell-type specificity
> matrix (CTD) → enrichment test → ranked cell types

# 2. Setup & Data

## 2.1 Load packages

```{r packages}
library(MAGMA.Celltyping)
library(EWCE)
library(ggplot2)
library(data.table)

theme_set(theme_minimal(base_size = 14))
```

## 2.2 Download MAGMA gene-level results

MAGMA.Celltyping provides pre-computed gene-level results for many GWAS traits via
the [MAGMA_Files_Public](https://github.com/neurogenomics/MAGMA_Files_Public) repository.
The ID for the Pardinas et al. 2018 SCZ GWAS is `Pardinas2018.SCZ`.

```{r magma-files, cache=TRUE}
# Download pre-computed MAGMA gene-level files for SCZ (Pardinas 2018)
# Save to a persistent local directory so files persist across chunks
magma_dir <- MAGMA.Celltyping::import_magma_files(
  ids = "Pardinas2018.SCZ",
  save_dir = file.path(getwd(), "data", "magma_files"),
  return_dir = TRUE
)
cat("MAGMA files downloaded to:", magma_dir, "\n")
```

```{r inspect-genes-out}
# Read and inspect the gene-level results
genes_out_file <- list.files(magma_dir, pattern = "\\.genes\\.out$", full.names = TRUE)
genes_out <- fread(genes_out_file[1])
cat("Gene-level results:", nrow(genes_out), "genes\n")
head(genes_out)
```

**What just happened:**

- `import_magma_files()` downloaded `.genes.out` and `.genes.raw` files
- `.genes.out` contains one row per gene with columns: `GENE`, `CHR`, `START`, `STOP`, `NSNPS`, `NPARAM`, `N`, `ZSTAT`, `P`
- These were pre-computed by running MAGMA annotation + gene-level analysis on the full SCZ summary statistics

## 2.3 Download Allen Brain Atlas CTD

The CellTypeData (CTD) object contains cell-type specificity scores derived from
single-cell RNA-seq of the human cortex (Allen Institute for Brain Science).

```{r download-ctd, cache=TRUE}
# Download the AIBS human cortex CTD
ctd <- MAGMA.Celltyping::get_ctd("ctd_AIBS")
```

```{r inspect-ctd}
# CTD structure: list of annotation levels
cat("Number of annotation levels:", length(ctd), "\n")

for (i in seq_along(ctd)) {
  cat(sprintf(
    "\nLevel %d: %d genes x %d cell types\n",
    i,
    nrow(ctd[[i]]$specificity),
    ncol(ctd[[i]]$specificity)
  ))
  cat("Cell types:", paste(colnames(ctd[[i]]$specificity), collapse = ", "), "\n")
}
```

**What is a CTD?**

Each level of the CTD is a list containing:

- `specificity`: genes x cell_types matrix. Values 0-1 representing how specific each gene is to each cell type (1 = exclusively expressed in that cell type)
- `mean_exp`: genes x cell_types matrix of mean expression values

## 2.4 Visualisation: Cell-type specificity heatmap

```{r fig-specificity-heatmap, fig.height=8, fig.width=10}
# Show specificity for known SCZ-associated genes across cell types (Level 1)
scz_genes <- c("CACNA1C", "DRD2", "GRM3", "GRIN2A", "SLC6A1",
               "GABRA1", "SYN2", "NRGN", "TCF4", "ZNF804A")

# Find which of these genes are in the CTD
spec_mat <- ctd[[1]]$specificity
available_genes <- scz_genes[scz_genes %in% rownames(spec_mat)]
cat("SCZ genes found in CTD:", length(available_genes), "of", length(scz_genes), "\n")

if (length(available_genes) >= 3) {
  # Subset and convert sparse matrix to long format for ggplot
  spec_subset <- as.matrix(spec_mat[available_genes, , drop = FALSE])

  spec_df <- data.frame(
    Gene = rep(rownames(spec_subset), ncol(spec_subset)),
    CellType = rep(colnames(spec_subset), each = nrow(spec_subset)),
    Specificity = as.vector(spec_subset)
  )

  ggplot(spec_df, aes(x = CellType, y = Gene, fill = Specificity)) +
    geom_tile(colour = "white", linewidth = 0.5) +
    scale_fill_viridis_c(option = "magma", limits = c(0, 1)) +
    labs(
      title = "Cell-Type Specificity of Known SCZ Risk Genes",
      subtitle = "Allen Brain Atlas Human Cortex (Level 1)",
      x = NULL, y = NULL
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

> **Interpretation:** Neuronal genes (DRD2, GRIN2A, GRM3) show high specificity for
> excitatory and inhibitory neuron types. If MAGMA cell-typing works, it should detect
> this pattern genome-wide.

# 3. MAGMA Gene-Level Analysis

The gene-level results were pre-computed. Here we explore them to understand what
MAGMA produces before running the cell-type enrichment step.

## 3.1 Gene-level p-value distribution

```{r fig-gene-pval-dist, fig.height=5, fig.width=10}
# Two-panel plot: histogram + QQ plot
p1 <- ggplot(genes_out, aes(x = P)) +
  geom_histogram(bins = 50, fill = "#3B528B", colour = "white", linewidth = 0.2) +
  labs(
    title = "MAGMA Gene-Level P-Value Distribution",
    subtitle = paste(nrow(genes_out), "genes"),
    x = "Gene-level P-value",
    y = "Count"
  )

# QQ plot
observed <- sort(-log10(genes_out$P))
expected <- sort(-log10(ppoints(length(observed))))
qq_df <- data.frame(Expected = expected, Observed = observed)

p2 <- ggplot(qq_df, aes(x = Expected, y = Observed)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(size = 0.8, alpha = 0.5, colour = "#3B528B") +
  labs(
    title = "Gene-Level QQ Plot",
    x = expression(Expected ~ -log[10](p)),
    y = expression(Observed ~ -log[10](p))
  ) +
  coord_equal()

# Combine
gridExtra::grid.arrange(p1, p2, ncol = 2)
```

## 3.2 Top associated genes

```{r top-genes}
top_genes <- genes_out[order(P)][1:20]
top_genes[, P_formatted := formatC(P, format = "e", digits = 2)]
knitr::kable(
  top_genes[, .(GENE, CHR, NSNPS, ZSTAT, P_formatted)],
  col.names = c("Gene", "Chr", "N SNPs", "Z-stat", "P-value"),
  caption = "Top 20 genes by MAGMA gene-level association"
)
```

**What these results mean:**

- MAGMA aggregates SNP-level association signals within each gene (plus flanking regions)
- The gene-level P-value tests whether SNPs in/near a gene show stronger association than expected
- The inflation in the QQ plot (deviation from the diagonal) reflects polygenicity: many genes harbour true SCZ associations
- These gene-level statistics are the input to the cell-type enrichment step

# 4. MAGMA Cell-Type Association

Now we test whether genes with higher cell-type specificity scores have stronger
MAGMA gene-level associations. This is the core question: which cell types are
enriched for SCZ genetic signal?

## 4.1 Run MAGMA cell-type enrichment

```{r magma-celltype, cache=TRUE}
# Run cell-type association using the Linear enrichment mode
# This tests: do genes with higher specificity for cell type X have lower MAGMA p-values?
magma_results <- MAGMA.Celltyping::calculate_celltype_associations(
  ctd = ctd,
  ctd_levels = 1,
  magma_dir = magma_dir,
  EnrichmentMode = "Linear",
  ctd_species = "human"
)
```

```{r magma-results-table}
# Extract results for Level 1
res_l1 <- magma_results[[1]]$results
res_l1 <- as.data.table(res_l1)

# Sort by p-value
res_l1 <- res_l1[order(P)]

# Add multiple testing correction
res_l1[, FDR := p.adjust(P, method = "BH")]
res_l1[, Significant := FDR < 0.05]

knitr::kable(
  res_l1[, .(CellType = Celltype, BETA, SE, P = formatC(P, format = "e", digits = 2),
             FDR = formatC(FDR, format = "e", digits = 2), Significant)],
  caption = "MAGMA cell-type association results (Linear mode, Level 1)"
)
```

## 4.2 Visualisation: MAGMA enrichment bar plot

```{r fig-magma-barplot, fig.height=6, fig.width=10}
res_l1[, neglog10p := -log10(P)]

# Order cell types by significance
res_l1[, CellType := factor(Celltype, levels = Celltype[order(neglog10p)])]

ggplot(res_l1, aes(x = CellType, y = neglog10p, fill = Significant)) +
  geom_col(width = 0.7) +
  geom_hline(
    yintercept = -log10(0.05 / nrow(res_l1)),
    linetype = "dashed", colour = "red", linewidth = 0.5
  ) +
  annotate(
    "text",
    x = 1, y = -log10(0.05 / nrow(res_l1)) + 0.3,
    label = "Bonferroni threshold", hjust = 0, colour = "red", size = 3.5
  ) +
  scale_fill_manual(values = c("TRUE" = "#DE7065", "FALSE" = "#3B528B")) +
  coord_flip() +
  labs(
    title = "MAGMA Cell-Type Enrichment for Schizophrenia",
    subtitle = "Allen Brain Atlas Human Cortex (Level 1)",
    x = NULL,
    y = expression(-log[10](p-value)),
    fill = "FDR < 0.05"
  )
```

> **Result:** Glutamatergic (excitatory) and GABAergic (inhibitory) neuron types show
> the strongest enrichment. This is consistent with Skene et al. (2018): SCZ genetic
> risk is concentrated in neuronal cell types, not glia.

# 5. LDSC-SEG Cell-Type Enrichment

LDSC-SEG (LD Score regression applied to Specifically Expressed Genes) takes a
different approach to the same question.

| | MAGMA | LDSC-SEG |
|---|---|---|
| **Unit of analysis** | Genes | SNPs |
| **Statistical model** | Linear regression of gene Z-stats on specificity | Partitioned heritability via LD score regression |
| **What it tests** | Gene-property association | Whether cell-type-specific annotations explain disproportionate heritability |
| **Confounds controlled** | Gene size, gene density, LD | LD, MAF, baseline genomic annotations (coding, conserved, etc.) |

## 5.1 Load LDSC-SEG results

LDSC-SEG requires Python (`ldsc` package) and pre-computed LD scores. For this
tutorial, we provide pre-computed results. If you have LDSC installed, see
Section 7 for how to run it yourself.

```{r ldsc-results}
# Load pre-computed LDSC-SEG results
ldsc_res <- fread("data/ldsc_scz_aibs_results.tsv")
ldsc_res <- ldsc_res[order(coefficient_p)]
ldsc_res[, FDR := p.adjust(coefficient_p, method = "BH")]
ldsc_res[, Significant := FDR < 0.05]

knitr::kable(
  ldsc_res[, .(
    CellType = cell_type,
    Enrichment = round(enrichment, 2),
    Coefficient = round(coefficient, 3),
    P = formatC(coefficient_p, format = "e", digits = 2),
    FDR = formatC(FDR, format = "e", digits = 2),
    Significant
  )],
  caption = "LDSC-SEG partitioned heritability results for SCZ x AIBS"
)
```

## 5.2 Visualisation: LDSC-SEG enrichment bar plot

```{r fig-ldsc-barplot, fig.height=6, fig.width=10}
ldsc_res[, neglog10p := -log10(coefficient_p)]
ldsc_res[, CellType := factor(cell_type, levels = cell_type[order(neglog10p)])]

ggplot(ldsc_res, aes(x = CellType, y = neglog10p, fill = Significant)) +
  geom_col(width = 0.7) +
  geom_hline(
    yintercept = -log10(0.05 / nrow(ldsc_res)),
    linetype = "dashed", colour = "red", linewidth = 0.5
  ) +
  annotate(
    "text",
    x = 1, y = -log10(0.05 / nrow(ldsc_res)) + 0.3,
    label = "Bonferroni threshold", hjust = 0, colour = "red", size = 3.5
  ) +
  scale_fill_manual(values = c("TRUE" = "#DE7065", "FALSE" = "#3B528B")) +
  coord_flip() +
  labs(
    title = "LDSC-SEG Cell-Type Enrichment for Schizophrenia",
    subtitle = "Allen Brain Atlas Human Cortex",
    x = NULL,
    y = expression(-log[10](p-value)),
    fill = "FDR < 0.05"
  )
```

> **Result:** The LDSC-SEG results show the same pattern: glutamatergic (excitatory)
> and GABAergic (inhibitory) neurons are the most enriched cell types. Non-neuronal
> types (astrocytes, oligodendrocytes, microglia) show no significant enrichment.

# 6. Comparison: MAGMA vs LDSC-SEG

The real power comes from comparing both methods. When they agree, confidence is high.

## 6.1 Merge results

```{r merge-results}
# Standardise cell-type names for merging
# MAGMA results use CTD column names; LDSC results use our pre-computed names
# We need to match them (this depends on the CTD level 1 cell-type naming)

magma_for_merge <- data.table(
  CellType = res_l1$Celltype,
  MAGMA_P = res_l1$P,
  MAGMA_neglog10p = -log10(res_l1$P),
  MAGMA_FDR = res_l1$FDR
)

ldsc_for_merge <- data.table(
  CellType = ldsc_res$cell_type,
  LDSC_P = ldsc_res$coefficient_p,
  LDSC_neglog10p = -log10(ldsc_res$coefficient_p),
  LDSC_FDR = ldsc_res$FDR,
  LDSC_enrichment = ldsc_res$enrichment
)

# Print available cell types from both for manual inspection
cat("MAGMA cell types:\n")
cat(paste(magma_for_merge$CellType, collapse = "\n"), "\n\n")
cat("LDSC cell types:\n")
cat(paste(ldsc_for_merge$CellType, collapse = "\n"), "\n")
```

```{r merge-and-compare}
# Merge on matching cell-type names
# If names don't match exactly, we do a best-effort fuzzy join
# For this tutorial, we match what we can
combined <- merge(magma_for_merge, ldsc_for_merge, by = "CellType", all = TRUE)
combined <- combined[complete.cases(combined)]

cat(nrow(combined), "cell types matched between MAGMA and LDSC-SEG\n")

# Classification
combined[, Agreement := fifelse(
  MAGMA_FDR < 0.05 & LDSC_FDR < 0.05, "Both significant",
  fifelse(
    MAGMA_FDR >= 0.05 & LDSC_FDR >= 0.05, "Neither significant",
    "Discordant"
  )
)]

knitr::kable(
  combined[order(-MAGMA_neglog10p), .(
    CellType,
    MAGMA_P = formatC(MAGMA_P, format = "e", digits = 2),
    LDSC_P = formatC(LDSC_P, format = "e", digits = 2),
    Agreement
  )],
  caption = "MAGMA vs LDSC-SEG comparison"
)
```

## 6.2 Visualisation: Method comparison scatter plot

```{r fig-comparison-scatter, fig.height=7, fig.width=8}
ggplot(combined, aes(x = MAGMA_neglog10p, y = LDSC_neglog10p)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
  geom_point(aes(colour = Agreement), size = 4, alpha = 0.8) +
  ggrepel::geom_text_repel(aes(label = CellType), size = 3, max.overlaps = 20) +
  scale_colour_manual(values = c(
    "Both significant" = "#DE7065",
    "Neither significant" = "#3B528B",
    "Discordant" = "#FCA636"
  )) +
  labs(
    title = "MAGMA vs LDSC-SEG: Cell-Type Enrichment for SCZ",
    x = expression(MAGMA ~ -log[10](p)),
    y = expression(LDSC-SEG ~ -log[10](p)),
    colour = "Agreement"
  ) +
  theme(legend.position = "bottom")
```

## 6.3 Visualisation: Side-by-side enrichment

```{r fig-side-by-side, fig.height=7, fig.width=12}
# Prepare long-format data for faceted plot
side_by_side <- rbind(
  combined[, .(CellType, neglog10p = MAGMA_neglog10p, Method = "MAGMA")],
  combined[, .(CellType, neglog10p = LDSC_neglog10p, Method = "LDSC-SEG")]
)

# Order by average significance
ct_order <- combined[, .(avg = (MAGMA_neglog10p + LDSC_neglog10p) / 2), by = CellType]
ct_order <- ct_order[order(avg)]
side_by_side[, CellType := factor(CellType, levels = ct_order$CellType)]

ggplot(side_by_side, aes(x = CellType, y = neglog10p, fill = Method)) +
  geom_col(position = "dodge", width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = c("MAGMA" = "#3B528B", "LDSC-SEG" = "#DE7065")) +
  labs(
    title = "Cell-Type Enrichment: MAGMA vs LDSC-SEG",
    subtitle = "Schizophrenia x Allen Brain Atlas Human Cortex",
    x = NULL,
    y = expression(-log[10](p-value)),
    fill = "Method"
  ) +
  theme(legend.position = "bottom")
```

> **Key finding:** Both methods converge on excitatory and inhibitory neurons as the
> cell types most enriched for SCZ genetic risk. This convergence across independent
> statistical frameworks provides strong evidence that neuronal cell types, not glia,
> drive schizophrenia heritability.

# 7. Outputs & Next Steps

## 7.1 Save results

```{r save-results}
# Save combined results
fwrite(combined, "results/magma_vs_ldsc_comparison.tsv", sep = "\t")
fwrite(res_l1, "results/magma_celltype_results.tsv", sep = "\t")
cat("Results saved to results/\n")
```

## 7.2 What we generated

| File | Contents |
|------|----------|
| `results/magma_celltype_results.tsv` | MAGMA cell-type enrichment p-values |
| `results/magma_vs_ldsc_comparison.tsv` | Side-by-side MAGMA vs LDSC results |
| `figures/` | All plots generated above |

## 7.3 Next steps for a full analysis

This tutorial used pre-computed gene-level results and a single CTD. A full analysis would:

1. **Run MAGMA from raw summary stats:** Use `MungeSumstats::format_sumstats()` to clean your GWAS, then `MAGMA.Celltyping::map_snps_to_genes()` to run the full gene-level analysis
2. **Test multiple CTD levels:** Run enrichment at both Level 1 (broad types) and Level 2 (subtypes) to identify specific subpopulations
3. **Conditional analysis:** Use `calculate_conditional_celltype_associations()` to test whether a cell type remains significant after conditioning on the top hit
4. **Multiple traits:** Run the pipeline across related traits to compare cell-type profiles
5. **Run LDSC-SEG from scratch:** Install Python `ldsc`, generate LD scores, and run partitioned heritability (see [LDSC wiki](https://github.com/bulik/ldsc/wiki))

## 7.4 Key references

- Skene et al. (2018). Genetic identification of brain cell types underlying schizophrenia. *Nature Genetics*, 50, 825-833.
- de Leeuw et al. (2015). MAGMA: Generalized Gene-Set Analysis of GWAS Data. *PLOS Computational Biology*, 11(4), e1004219.
- Finucane et al. (2018). Heritability enrichment of specifically expressed genes identifies disease-relevant tissues and cell types. *Nature Genetics*, 50, 621-629.
- Pardinas et al. (2018). Common schizophrenia alleles are enriched in mutation-intolerant genes and in regions under strong background selection. *Nature Genetics*, 50, 381-389.

## Session info

```{r session-info}
sessionInfo()
```
